---
title: "R Text Wrangling Basics"
author: Hauke Licht
institute: University of Innsbruck
date: last-modified
date-format: "MMMM D, YYYY"
toc: true
toc-depth: 2
toc-location: right
toc-expand: true
execute:
  echo: true
  eval: true
  message: false
  warning: false
code-annotations: select
---

# Recap: Types of vectors

To understand how text data is handled in R, it's important to remember some R programming basics:

#### Terminology

- **vector**: a sequence of values
- **element**/**value**: a single value in a vector

```{r}
#| eval: false
# creating a numeric vector called "vec"
vec <- c(1, 2, 3)
val <- vec[1]
```

#### Vector Types in R

In R, there are four main types of vectors: 

1. `logical` for "boolean" (true/false) values
2. `integer` for whole numbers (e.g., ... -99, ... -1, 0, 1, 2, ...) 
3. `double` for real-valued numbers (e.g., 1.3)
4. **`character`** for representing **text**

Each type has its own characteristics and is used for different purposes.


#### `NA` values

::: {.callout-note icon=false}
`NA` is the special value used in R to represent missing or undefined values
:::


There are special `NA` values for each vector type:

1. logical &rarr; `NA`
2. integer &rarr; `NA_integer_`
3. double &rarr; `NA_real_` (don't ask me why it's not `NA_double_` 😅)
4. **character** &rarr; **`NA_character_`**

```{r}
typeof(NA)
typeof(as.character(NA))
typeof(NA_character_)
```


# Basic R functions for text wrangling

::: {.callout-note icon=false}
**Text wrangling** is a term  I use for referring to anything I do with a computer to process textual data.
:::


We'll look at **three groups** of text wrangling functions:

1. Functions for *creating* character vectors
2. Functions for *manipulating* ("changing") character vectors
3. Functions for *analyzing* character vectors

## Functions for _creating_ character vectors

### c()

`c()` combines values into a vector

#### Example

```{r}
x <- c("Hello", "World")
x
```


### rep()

`rep()` replicates values in a vector

#### Arguments

- `x`: Value to be replicated
- `times`: Number of times to replicate the value
- `each`: Number of times to repeat each value

#### Examples

```{r}
x <- rep("Hello", times = 3)
x
```

```{r}
rep(c("Hello", "World"), each = 2)
rep(c("Hello", "World"), times = 2)
```


### paste()

`paste()` combines or concatenates several character values into a single character value

#### Arguments

- `...`: Character vectors to be combined
- `sep`: Separator between the values (default: " ")
- `collapse`: Separator between the combined values (default: NULL)

#### Example

```{r}
x <- c("a", "b", "c")
y <- c("1", "2", "3")
paste(x, y, sep = ":")
```


```{r}
x <- c("a", "b", "c")
paste(x, collapse = ", ")
```



### sprintf()

`sprintf()` formats character values according to a specified format

#### Arguments

- `fmt`: Format string
- `...`: Values to be formatted

#### Example

```{r}
sprintf(fmt = "Hello, %s! My name is %s", "friends", "Hauke")
```

*Note:* we use `%s` as a placeholder to insert a character value


#### Formatting options

the most important formatting options (a.k.a "placeholders") are:

- `%s` for inserting a character value
- `%d` for inserting an integer value
- `%f` for inserting a double value

See <https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/sprintf> for more options

#### Example

`sprintf()` is great for creating identifiers (IDs):

```{r}
df <- data.frame(
  document_number = c(1, 1, 2, 2),
  paragraph_number = c(1, 2, 1, 2)
)
df$paragraph_id <- sprintf("d%02d_p%02d", df$document_number, df$paragraph_number)
df
```

*Note:* `%02d` is a placeholder for a two-digit integer with leading zeros

```{r}
sprintf("%02d", 8:12)
sprintf("%03d", c(1, 10, 100))
```


## Functions for _analyzing_ character vectors

### nchar()

`nchar()` counts the number of characters in a character value

#### Arguments

- `x`: Character vector

#### Example

```{r}
x <- "Hello, World!"
nchar(x)
```

### grepl()

`grepl()` tests if a pattern is present in a character vector

#### Arguments

- `pattern`: Pattern to be matched
- `x`: Character vector

#### Example

```{r}
x <- "Hello, World!"
grepl("Hello", x)
grepl("banana", x)
grepl("hello", x, ignore.case = TRUE)
```

*Note:* be careful with `NA` values:

```{r}
grepl(pattern = "Hello", x = NA)
grepl(pattern = NA, x = "Hello")
```


### grep()

`grep()` returns the indices of the values in a character vector that contain a pattern

#### Arguments

- `pattern`: Pattern to be matched
- `x`: Character vector
- `value`: Logical value (`TRUE`/`FALSE`) indicating whether to return the values instead of the indices

#### Example

```{r}

x <- c("apple", "banana", "cherry")
grep("a", x)
grep("a", x, value = TRUE)
```


## Functions for _manipulating_ character vectors

### `trimws`

`trimws()` removes leading and trailing white space from a character value

#### Arguments

- `x`: Character vector

#### Example

```{r}
x <- "  Hello, World!  "
trimws(x)
```


### `tolower()` and `toupper()`

`tolower()` and `toupper()` convert text to lowercase and uppercase, respectively

#### Arguments

- `x`: Character vector

#### Example

```{r}
x <- "Hello, World!"
tolower(x)
toupper(x)
```


### `substr()`

`substr()` extracts a substring from a character value

#### Arguments

- `x`: Character vector
- `start`: Starting position of the substring
- `stop`: Ending position of the substring

#### Example

```{r}
x <- "Hello, World!"
substr(x, start = 1, stop = 5)
```


### `sub` and `gsub`

- `sub()` replaces the first occurrence of a pattern in a character value
- `gsub()` replaces all occurrences of a pattern in a character value

#### Arguments

- `pattern`: Pattern to be replaced
- `replacement`: Replacement value
- `x`: Character vector

#### Example

```{r}
x <- "Hello, World!"
sub(pattern = "World", replacement = "Universe", x)
gsub(pattern = "o", replacement = "🤗", x)
```


### strsplit

`strsplit()` splits a character value into substrings based on a specified delimiter

#### Arguments

- `x`: Character vector
- `split`: Delimiter to split the character value

#### Example

```{r}
x <- "Hello, World!"
strsplit(x, split = ", ")
```

**_Important:_** `strsplit()` returns a **list** of character vectors


---

# `stringr`

## The `stringr` package

- The `stringr` package provides a set of functions for working with character vectors
- The functions in `stringr` are designed to be more consistent and easier to use than the base R functions

#### How to

```{r}
library(stringr)
```

*Note:* `stringr` is part of the "tidyverse" and is loaded automatically when you load the `tidyverse` package


## `stringr` equivalents to base R functions

- `str_c()` is equivalent to `paste()`
- `str_length()` is equivalent to `nchar()`
- `str_to_lower()` and `str_to_upper()` are equivalent to `tolower()` and `toupper()`
- `str_sub()` is equivalent to `substr()`
- `str_replace()` and `str_replace_all()` are equivalent to `sub()` and `gsub()`
- `str_split()` is equivalent to `strsplit()`
- `str_detect()` is equivalent to `grepl()`
- `str_locate()` is equivalent to `grep()`


**Practical advantage:** `stringr` functions are defined and named more consistently

- the first argument is always the character vector
- if applicable, the second argument is always the pattern
- if applicable, the third argument is always the replacement value

#### Examples

```{r}
str_replace("Hello, World!", pattern = "World", replacement = "Universe")
# instead of `sub(World, "Universe", "Hello, World!")`

str_detect("Hello, World!", "Hello")
# instead of `grepl("Hello", "Hello, World!")`

str_locate("Hello, World!", "World")
# instead of `grep("World", "Hello, World!")`
```


## Other useful stringr functions

- `str_trim()`: removes leading and trailing whitespace
- `str_pad()`: pads a string with spaces
- `str_wrap()`: wraps a string to a specified width
- `str_sort()`: sorts a character vector
- `str_order()`: returns the order of a character vector
- `str_replace_na()`: replaces `NA` values with a specified value
- `str_extract()`: extracts a pattern from a character vector

