---
title: "Replication of Kozlowski *et al.* \"The Geometry of Culture: Analyzing the Meanings of Class through Word Embeddings\""
author: Hauke Licht
institute: University of Innsbruck
date: last-modified
date-format: "MMMM D, YYYY"
toc: true
toc-depth: 2
toc-location: right
toc-expand: true
execute:
  echo: true
  eval: true
  message: false
  warning: false
  cache: true
code-annotations: select
format:
  html:
    embed-resources: true
    self-contained: true
    anchor-sections: true
    smooth-scroll: true
---

In their much-cited [publication](https://journals.sagepub.com/doi/full/10.1177/0003122419877135) "The Geometry of Culture: Analyzing the Meanings of Class through Word Embeddings," Kozlowski, Taddy, and Evans show that conceptual dimensions in the embedding spaces correspond to dimensions of cultural meaning, and words' relation with these dimensions reflects widely shared cultural associations.

## Replication: goals and approach

In this notebook, we replicate parts of their analyses.
Our goal is to see if we find the same patterns they describe in their publication.
We do this using their original word lists but a different word embedding model.


## Setup

```{r}
#| message: false
#| warnings: false
library(readr)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)

base_path <- file.path("..", "..") # NOTE only when running in quarto notebook
data_path <- file.path(base_path, "data", "replications", "kozlowski_geometry_2019")
```

::: {.callout-note title="Model download"}

I'm using the 200D version of Stanford NLP's GloVe model here instead of the 50D version we used in the word embedding intro notebook.
Go to script `models/download_glove_vectors.R` to download it if you don't have it yet _or_ replace 
`model_file <- "glove.2024.wikigiga.200d.zip"` below with `model_file <- "glove.2024.wikigiga.50d.zip"`.
Note that if you choose the latter option, you'll get different results than in the pre-compiled notebook.

:::

```{r}
#| cache: true
# Load pre-trained GloVe word embeddings
models_path <- file.path(base_path, "models")
# model_file <- "glove.2024.wikigiga.50d.zip"
model_file <- "glove.2024.wikigiga.200d.zip"
fp <- file.path(models_path, model_file)

# read the vectors from the zip file
wv <- data.table::fread(
  fp,
  skip = 1, # skip header line
  header = FALSE,
  sep = " ", # words and vector values are separated by white space
  quote = "", # no quote characters
  data.table = FALSE # return as data frame
)

# set row names to words and remove word column
row.names(wv) <- wv[[1]]
wv[[1]] <- NULL
names(wv) <- paste0("V", seq_len(ncol(wv)))
wv <- as.matrix(wv)
```

## Kozlowski *et al.* measurement strategy


#### *1. Inducing dimensions*

Kozlowski _et al._ explain that their technique for inducing cultural dimensions in the word embedding vector space builds on logic of **solving analogy tasks**:

- Adding the vector $(v_{\text{affluence}}−v_{\text{poverty}})$ to a word's embedding is equivalent to taking one "step" in the direction of affluence.
- To tap more broadly into this latent "affluence" dimension, you can do the same computation for more polar word pairs than just ("affluence", "poverty"), such as 
such as ("rich", "poor"), ("priceless", "worthless"), or ("prosperous", "bankrupt").
- Averaging the differences of several related antonym pairs then yields "a single, robust affluence dimension".

#### *2. Computing words' associations*

To identify the cultural associations/valence of a word, they compute the cosine similarity of its normalized vector with a given cultural dimension.

*Note:* They note that this is similar to "calculate the orthogonal projection of the word vector onto the cultural dimension of interest."

**_Example_**

Say the class dimensions is constructed from the polar word pairs  ("affluence", "poverty") , ("rich", "poor"), ("priceless", "worthless"), etc.
Then a more positive cosine value would indicate association with affluence, and more negative values an association with poverty.

#### Summary of ingredients

1. a pre-trained word embedding
2. a set of polar word (i.e., antonym) pairs for each cultural dimension
3. a bunch of words that should be related to ("projected onto") these dimensions

## Implementing their analyses with R

### 1. Load the antonym words pairs

I have prepared the data in their category-specific antonym word pairs lists into a long tab-separated file:

```{r}
#| cache: true
fp <- file.path(data_path, "antonyms.tsv")
df <- read_tsv(fp, col_names = c("dimension", "pair"))
# split column pair into separate columns
df <- tidyr::separate(df, pair, into = c("word1", "word2"), sep = "-", remove = FALSE)
sample_n(df, 10)
```

```{r}
table(df$dimension)
```

### 2. Inducing a dimensions using antonym pairs 

Kozlowski's approach is to 

1. span vectors between the polar terms in each word pair (by subtracting the second term's embedding from the first one's) , and
2. then average these 'direction' vectors into one vector

They do this for *each* dimension separately!

```{r}
unique(df$dimension)
```

```{r}
# Look at first group as example
expl <- filter(df, dimension == "Affluence")
expl
```

```{r}
# Helper function to get word vector
get_word_vector <- function(word) {
  if (word %in% rownames(wv)) {
    return(wv[word, ])
  } else {
    return(NULL)
  }
}

# Test with first row
r <- expl[1, ]
word1_exists <- r$word1 %in% rownames(wv)
word2_exists <- r$word2 %in% rownames(wv)
cat("Both words exist in vocabulary:", word1_exists && word2_exists, "\n")
```

```{r}
if (word1_exists && word2_exists) {
  diff <- wv[r$word1, ] - wv[r$word2, ]
  cat("Difference vector shape:", length(diff), "\n")
}
```

```{r}
# Initialize empty list for dimensions
dims <- list()

# Group by dimension and compute average difference vectors
for (dim_name in unique(df$dimension)) {
  group <- filter(df, dimension == dim_name)
  
  # List to record each antonym pair's difference vector 
  diffs <- matrix(nrow = nrow(group), ncol = ncol(wv))
  colnames(diffs) <- colnames(wv)
  
  # Iterate over antonym pairs
  for (i in 1:nrow(group)) {
    r <- group[i, ]
    # Check if both words are in vocabulary
    if (r$word1 %in% rownames(wv) && r$word2 %in% rownames(wv)) {
      # If so, subtract them
      diffs[i, ] <- wv[r$word1, ] - wv[r$word2, ]
    }
  }
  
  # Average antonym pairs' difference vectors (across rows)
  dims[[dim_name]] <- colMeans(diffs, na.rm = TRUE)
}
```

Now we have _one_ vector per cultural dimension that points from one set of polar terms to the opposite of polar terms for each dimension.

```{r}
names(dims)
```


### 3. Computing associations of words with cultural dimensions 

**_Note_** to perform the following computations, we need to normalize vectors. 
Here is a helper function to do this:

```{r}
# Normalizing the vectors
normalize_vector <- function(x) {
  return(x / sqrt(sum(x^2)))
}
```

::: {.callout-info title="Normalization"}

"Normalization" means scaling a vector so that its length (or magnitude) is 1.
This makes it easier to compare vectors based on their direction alone, without being influenced by their original lengths.

:::

#### Example with the 'gender' dimension

Let's check whether the occupation names "manager" and "secretary" are associated differently with our gender dimension:

```{r}
filter(df, dimension == "Gender")
```

```{r}
manager <- normalize_vector(wv["manager", ])
nurse <- normalize_vector(wv["nurse", ])

# Cosine similarity function
cosine_similarity <- function(a, b) {
  # Ensure both are vectors
  if (is.matrix(a)) a <- as.vector(a)
  if (is.matrix(b)) b <- as.vector(b)
  
  # Compute dot product divided by product of norms
  return(sum(a * b) / (sqrt(sum(a^2)) * sqrt(sum(b^2))))
}

cat("manager:", cosine_similarity(manager, normalize_vector(dims[["Gender"]])), "\n")
cat("nurse:", cosine_similarity(nurse, normalize_vector(dims[["Gender"]])), "\n")

cat("\nfor comparison:\n")
cat("man:", cosine_similarity(normalize_vector(wv["man", ]), normalize_vector(dims[["Gender"]])), "\n")
cat("woman:", cosine_similarity(normalize_vector(wv["woman", ]), normalize_vector(dims[["Gender"]])), "\n")
```

**Interpretation:** 

- "Manager" is more associated with the 'male' pole of the gender dimension.
- "Nurse" is more associated with the 'female' pole of the gender dimension.
- the `for comparison` numbers show how the terms "man" and "women" are associated with the 'male' and 'female' poles of the gender dimension

#### Implementation with all terms and the "affluence" dimensions

Let's load the terms

```{r}
#| cache: true
fp <- file.path(data_path, "cultural_association_words.tsv")
df_words <- read_tsv(fp, col_types = cols())
table(df_words$category)
```

```{r}
sample_n(df_words, 10)
```

Because these terms are likely to have been used more in their lowercase versions, we'll lowercase them:

```{r}
df_words$word <- tolower(df_words$word)
```

Now, let's get the list of categories in `df_words` and compute scores for the "Affluence" dimension of culture:

```{r}
# Example with the "affluence" dimension
dim_name <- "Affluence"
scores <- list()

# Normalize dimension vectors
dims_normalized <- lapply(dims, normalize_vector)

for (i in 1:nrow(df_words)) {
  r <- df_words[i, ]
  if (!r$word %in% rownames(wv))
    next
  
  scores[[r$word]] <- r
  v <- normalize_vector(wv[r$word, ])
  scores[[r$word]]$score <- cosine_similarity(v, dims_normalized[[dim_name]])
}
scores_df <- bind_rows(scores)
```

```{r}
# Just show results for words in the "Clothing" category 
scores_df |> 
  filter(category == "Clothing") |> 
  arrange(desc(score))
```

Let's visualize terms associations with the Affluence dimension (by category):

```{r}
#| fig-height: 15

# Create a basic scatter plot
ggplot(scores_df, aes(x = score, y = reorder(word, desc(score)))) +
  geom_col(color = NA) +
  facet_grid(rows = vars(category), scales = "free_y", space = "free") +
  labs(x = "Association with \"Affluence\" dimension", y = NULL) +
  theme_minimal()
```

### Replicate their Figure 3

```{r}
words <- c("jazz", "rap", "opera", "punk", "techno", "hip_hop", "bluegrass")

associations <- list()
# Race associations (see Kozlowski et al., 2019, Figure 3)
for (dim_name in c("Affluence", "Race")) {
  associations[[dim_name]] <- list()
  for (w in words) {
    if (w %in% rownames(wv)) {
      v <- normalize_vector(wv[w, ])
      cs <- cosine_similarity(v, dims_normalized[[dim_name]])
      associations[[dim_name]][[w]] <- cs
    }
  }
}
```

```{r}
# Convert to data frame for plotting
plot_data <- data.frame(
  word = names(associations[["Affluence"]]),
  Race = unlist(associations[["Race"]]),
  Affluence = unlist(associations[["Affluence"]])
)

# Create scatterplot
ggplot(plot_data, aes(x = Race, y = Affluence)) +
  geom_point(color = "red", alpha = 0.4) +
  geom_text(aes(label = word), vjust = -0.5, hjust = 0.5) +
  labs(x = "Race", y = "Affluence") +
  theme_minimal()
```

### Replicate their Figure 5

```{r}
# Compute similarities between all dimension pairs
dim_names <- names(dims_normalized)
dim_sims <- matrix(NA, nrow = length(dim_names), ncol = length(dim_names))
rownames(dim_sims) <- dim_names
colnames(dim_sims) <- dim_names

for (i in 1:length(dim_names)) {
  for (j in 1:length(dim_names)) {
    if (i != j) {
      dim_sims[i, j] <- cosine_similarity(
        dims_normalized[[dim_names[i]]], 
        dims_normalized[[dim_names[j]]]
      )
    }
  }
}
```

```{r}
# Convert matrix to long format for ggplot
dim_sims_df <- data.frame(
  Dimension1 = rep(rownames(dim_sims), times = ncol(dim_sims)),
  Dimension2 = rep(colnames(dim_sims), each = nrow(dim_sims)),
  Similarity = as.numeric(dim_sims)
)
dim_sims_df <- filter(dim_sims_df, !is.na(Similarity))
# melt(dim_sims, na.rm = TRUE)
# names(dim_sims_df) <- c("Dimension1", "Dimension2", "Similarity")

# Create heatmap
ggplot(dim_sims_df, aes(x = Dimension2, y = Dimension1, fill = Similarity)) +
  geom_tile() +
  geom_text(aes(label = round(Similarity, 3)), color = "black", size = 3) +
  scale_fill_gradient2(low = "red", high = "green", mid = "yellow", midpoint = 0) +
  labs(title = "Dimension Similarities", x = "", y = "", fill = "Similarity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**_Note:_** Kozlowski _et al._ (2019) report the following similarities of other dimensions with the **Affluence** (class) dimension in Figure 5 for the period 1990:

- Cultivation: 0.31
- Education: 0.42
- Employment: 0.08
- Gender: -.05
- Morality: 0.1
- Status: 0.23

Our estimates are super similar!