---
title: "Replication of Caliskan *et al.* \"Semantics derived automatically from language corpora contain human-like biases\""
author: Hauke Licht
institute: University of Innsbruck
date: last-modified
date-format: "MMMM D, YYYY"
toc: true
toc-depth: 2
toc-location: right
toc-expand: true
execute:
  echo: true
  eval: true
  message: false
  warning: false
  cache: true
code-annotations: select
format:
  html:
    embed-resources: true
    self-contained: true
    anchor-sections: true
    smooth-scroll: true
---

In their often-cited *Science* [publication](https://www.science.org/doi/10.1126/science.aal4230) "Semantics derived automatically from language corpora contain human-like biases," Caliskan, Bryson, and Narayanan propose a method for quantifying the biases captured in word embedding models.

In their abstract, they write:

> ... we show that applying machine learning to ordinary human language results in human-like semantic biases. 
> We replicate a spectrum of known biases, as measured by the *Implicit Association Test*, using a widely used, purely statistical machine-learning model trained on a standard corpus of text from the Web.
> Our results indicate that text corpora contain recoverable and accurate imprints of our historic biases, whether morally neutral as towards insects or flowers, problematic as towards race or gender, or even simply veridical, reflecting the status quo distribution of gender with respect to careers or first names.


## Replication: goals and approach

In this notebook, we replicate their analyses.
Our goal is to see if we find the same patterns they describe in their publication.
We do this using their original word lists but a different word embedding model.

Our replication will focus on their **WEAT** metric – the *Word Embedding Association Test*.


## Setup

Before we can get going, let's set up our notebook:

```{r}
library(jsonlite)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
base_path <- file.path("..", "..") # NOTE only when running in quarto notebook
data_path <-  file.path(base_path, "data", "replications", "caliskan_semantics_2017")
```

**_Note:_** 
We'll work with data collected by Github user `chadaeun` for his "weat_replication" project.
The relevant data is available at https://github.com/chadaeun/weat_replication/blob/master/weat/weat.json and has already been downloaded

## The WEAT

Let

- $X$ and $Y$ be two sets of **_target words_** of equal size (to be tested for association), and
- $A$ and $B$ the two sets of **_attribute words_** (indicating conceptual opposites).

#### Example

- The *target words* could be occupations ("programmer", "engineer", "scientist"; and "nurse", "teacher", "librarian").
- The two sets of *attribute words* could be ("man", "male") and ("woman", "female").

#### Formula

The test statistic $s$ of the word-embedding association test (WEAT) is defined as 

$$
s(X, Y, A, B)=\sum_{x \in X} s(x, A, B)-\sum_{y \in Y} s(y, A, B)
$$

where 

$$
s(w, A, B) =\operatorname{mean}_{a \in A} \cos (\vec{w}, \vec{a})-\operatorname{mean}_{b \in B} \cos (\vec{w}, \vec{b})
$$

measures the association of $w$ with the attribute, and
$s(X,Y,A,B)$ measures the differential association of the two sets of target words $A$ and $B$ with the attribute.

The **null hypothesis** is that there is no difference between the two sets of target words in terms of their relative similarity to the two sets of attribute words, i.e., $\text{H}_0: s = 0$

## Example with code

### Load pre-trained GloVe word embeddings

::: {.callout-note title="Model download"}

I'm using the 200D version of Stanford NLP's GloVe model here instead of the 50D version we used in the word embedding intro notebook.
Go to script `models/download_glove_vectors.R` to download it if you don't have it yet _or_ replace 
`model_file <- "glove.2024.wikigiga.200d.zip"` below with `model_file <- "glove.2024.wikigiga.50d.zip"`.
Note that if you choose the latter option, you'll get different results than in the pre-compiled notebook.

:::


```{r}
#| cache: true
models_path <- file.path(base_path, "models")
# model_file <- "glove.2024.wikigiga.50d.zip"
model_file <- "glove.2024.wikigiga.200d.zip"
fp <- file.path(models_path, model_file)

# read the vectors from the zip file
wv <- data.table::fread(
  fp,
  skip = 1, # skip header line
  header = FALSE,
  sep = " ", # words and vector values are separated by white space
  quote = "", # no quote characters
  data.table = FALSE # return as data frame
)

# set row names to words and remove word column
row.names(wv) <- wv[[1]]
wv[[1]] <- NULL
names(wv) <- paste0("V", seq_len(ncol(wv)))
wv <- as.matrix(wv)
```

### Get the lists of target and attribute words

Let's first load the dictionary of wordlists Caliskan *et al.* (2017) used:

```{r}
fp <- file.path(data_path, "wordlists.json")
wordlists <- fromJSON(fp)
```

```{r}
names(wordlists)
```

```{r}
#| echo: false
# drop selected
wordlists["EuropeanAmerican_AfricanAmerican_Pleasant_Unpleasant_2"] <- NULL
```


Let's start with the example of the WEAT for science and art attributes with target words representing the concepts "male" and "female".

```{r}
data_dict <- wordlists[["Science_Arts_Male_Female"]]
names(data_dict)
```

**_Note:_** Each top-level dictionary element contains a dictionary with the same set of keys.

We want to use the following attribute words list for A and B:

```{r}
A_key <- data_dict$A_key
cat("A:", data_dict[[A_key]], "\n")
B_key <- data_dict$B_key
cat("B:", data_dict[[B_key]], "\n")
```

And for the target words X and Y, we use the following word lists:

```{r}
X_key <- data_dict$X_key
cat("X:", data_dict[[X_key]], "\n")
Y_key <- data_dict$Y_key
cat("Y:", data_dict[[Y_key]], "\n")
```

### Load word embeddings and get target and attribute words' embeddings

Next, we need to get these words embeddings.
We'll use our loaded GloVe model and use a helper function to extract word vectors from it:

```{r}
get_word_vectors <- function(words) {
  # Filter words that exist in the vocabulary
  words <- words[words %in% rownames(wv)]
  if (length(words) == 0) {
    stop("No words found in vocabulary")
  }
  # Return the word vectors as a matrix
  return(wv[words, , drop = FALSE])
}
```

```{r}
# test
tmp <- get_word_vectors(c("hello", "world"))
dim(tmp)
```

Now we can get the word vectors for words in A, B, X, and Y.

```{r}
A <- get_word_vectors(data_dict[[A_key]])
B <- get_word_vectors(data_dict[[B_key]])
X <- get_word_vectors(data_dict[[X_key]])
Y <- get_word_vectors(data_dict[[Y_key]])
```

It's important that A and B, and X and Y have the same numbers of rows!

```{r}
stopifnot(nrow(A) == nrow(B))
stopifnot(nrow(X) == nrow(Y))
```

### Computing association scores

Now we can compute 

$$
s(w, A, B) = \operatorname{mean}_{a \in A} \cos (\vec{w}, \vec{a})-\operatorname{mean}_{b \in B} \cos (\vec{w}, \vec{b})
$$

for all $\vec{w} \in \mathbf{X}$ and all $\vec{w} \in \mathbf{Y}$, respectively.

For this, we'll need **two helper functions**: 

1. one that normalizes the word vectors to unit vectors, and 
2. another that can compute the cosine similarity between two matrices. 

I have already implemented them.

```{r}
norm_vec <- function(vec) {
  if (is.matrix(vec)) {
    return(vec / sqrt(rowSums(vec^2)))
  } else {
    return(vec / sqrt(sum(vec^2)))
  }
}

cos_sim <- function(v1, v2) {
  # Ensure both inputs are matrices
  if (is.vector(v1)) v1 <- matrix(v1, nrow = 1)
  if (is.vector(v2)) v2 <- matrix(v2, nrow = 1)
  
  # Normalize vectors
  v1_norm <- norm_vec(v1)
  v2_norm <- norm_vec(v2)
  
  # Compute cosine similarity
  sim <- v1_norm %*% t(v2_norm)
  
  # Clip values to [-1, 1]
  sim <- pmax(pmin(sim, 1.0), -1.0)
  
  return(sim)
}
```

```{r}
dim(A)
```

Let's illustrate how `cos_sim()` works for $\mathbf{A}$ &mdash; our (8, 50) the matrix of embeddings from the list of "Male attributes".

We'll start with only the first row vector in $\mathbf{X}$:

```{r}
# get first vector in X
w <- X[1, , drop = FALSE]
# compute w's similarity with each row-vector in A
cos_sim(w, A)
```

Since `A` has 8 rows, we get 8 similarity scores.
For example, the first element is the cosine similarity between `w` (representing the 'science' dimension) and the first vector in `A` (representing term "brother").

But to compute $\operatorname{mean}_{a \in A} \cos (\vec{w}, \vec{a})$ for `w`, we want to *average* these similarities:

```{r}
mean(cos_sim(w, A))
```

To get the first element in the WEAT formula, we compute 

$$
s(w, A, B) =\operatorname{mean}_{a \in A} \cos (\vec{w}, \vec{a})-\operatorname{mean}_{b \in B} \cos (\vec{w}, \vec{b})
$$

In code, this is just:

```{r}
mean(cos_sim(w, A)) - mean(cos_sim(w, B))
```

The difference is positive, because `w` is, on average, slightly more similar to terms in `A` than to terms in `B`.

But, of course, we want to compute these quantities for each vector in `X` and `Y`, respectively.

Our `cos_sim()` function is able to handle this case:

```{r}
dim(cos_sim(X, A))
```

Here we have 

- eight rows, one for each term in **X**, and 
- eight columns, one for each term in **A**.

So to get one average similarity score per term in **X** we need to compute **_row averages_** (like `rowMeans` in R):

```{r}
rowMeans(cos_sim(X, A)) # <== summarize over columns (i.e. at row level)
```

So to get to the final _list of association scores_, we compute:

```{r}
rowMeans(cos_sim(X, A)) - rowMeans(cos_sim(X, B))
```

Let's define a custom function that does just that for input matrices `W`, `A`, and `B`:

```{r}
# from https://github.com/chadaeun/weat_replication/blob/0753713a47333827ef9f653d85e08740834ef698/lib/weat.py#L21C3-L21C3
weat_association <- function(W, A, B) {
  # Returns association of the word w in W with the attribute for WEAT score.
  # s(w, A, B)
  # :param W: target words' vector representations
  # :param A: attribute words' vector representations
  # :param B: attribute words' vector representations
  # :return: vector with length nrow(W). each element represents association of the word w in W
  return(rowMeans(cos_sim(W, A)) - rowMeans(cos_sim(W, B)))
}
```

### Computing the differential association score

Finally, we want to get from

$$
s(w, A, B) =\operatorname{mean}_{a \in A} \cos (\vec{w}, \vec{a})-\operatorname{mean}_{b \in B} \cos (\vec{w}, \vec{b})
$$

to the **differential association** score:

$$
s(X, Y, A, B)=\sum_{x \in X} s(x, A, B)-\sum_{y \in Y} s(y, A, B)
$$

To this end, we need to sum the the outputs of `weat_association` for both `X` and `Y` and subtract them:

```{r}
sum(weat_association(X, A, B)) - sum(weat_association(Y, A, B))
```

Let's wrap this last line of code in a function:

```{r}
# from https://github.com/chadaeun/weat_replication/blob/0753713a47333827ef9f653d85e08740834ef698/lib/weat.py#L33C1-L43C81
weat_differential_association <- function(X, Y, A, B) {
  # Returns differential association of two sets of target words with the attribute for WEAT score.
  # s(X, Y, A, B)
  # :param X: target words' vector representations
  # :param Y: target words' vector representations
  # :param A: attribute words' vector representations
  # :param B: attribute words' vector representations
  # :return: differential association (float value)
  return(sum(weat_association(X, A, B)) - sum(weat_association(Y, A, B)))
}
```

Recall what our `A`, `B`, `X`, and `Y` terms are:

```{r}
cat("A:", data_dict[[A_key]], "\n")
cat("B:", data_dict[[B_key]], "\n")
cat("X:", data_dict[[X_key]], "\n")
cat("Y:", data_dict[[Y_key]], "\n")
```

```{r}
weat_differential_association(X, Y, A, B)
```

**Interpretation:** 
The fact that the differential association score is positive indicates that, taken together, the science target words are on average more associated with male than female terms.

# Advanced

::: {.callout-info title="Advanced computations"}

The above sections are enough to understand the WEAT.
The computations covered below are advanced but required to replicate the authors' main finding, however.

:::

The sections below cover the computation of the **effect size** for the WEAT, which is a normalized measure of the strength of the association difference.

### Computing the effect size

But to get at the **effect size** for the WEAT (the "normalized measure of how separated the distributions of associations between the target and attribute"), we need to compute

$$
\frac{\operatorname{mean}_{x \in X} s(x, A, B)-\operatorname{mean}_{y \in Y} s(y, A, B)}{\operatorname{std} \_\operatorname{dev}_{w \in X \cup Y} s(w, A, B)}
$$

```{r}
x_association <- weat_association(X, A, B)
y_association <- weat_association(Y, A, B)
tmp1 <- mean(x_association) - mean(y_association)
tmp2 <- sd(c(x_association, y_association)) # <== the "union" of X and Y is just the concatenation of the two

effect_size <- tmp1/tmp2
effect_size
```

Let's wrap this in a function as well:

```{r}
weat_score <- function(X, Y, A, B) {
  # Returns WEAT score
  # X, Y, A, B must be matrices with rows representing words and columns representing dimensions
  # CAUTION: this function assumes that there's no intersection word between X and Y
  # :param X: target words' vector representations
  # :param Y: target words' vector representations
  # :param A: attribute words' vector representations
  # :param B: attribute words' vector representations
  # :return: WEAT score
  
  x_association <- weat_association(X, A, B)
  y_association <- weat_association(Y, A, B)

  tmp1 <- mean(x_association) - mean(y_association)
  tmp2 <- sd(c(x_association, y_association))

  return(tmp1 / tmp2)
}
```

## Complete replication: Get WEAT scores for different target and attribute term pairings

Our `wordlists` object contains more sets of words to compute the WEAT:

```{r}
names(wordlists)[sapply(wordlists, function(x) x$method == "weat")]
```

So let's loop over these tests to compute WEAT effect sizes for each.
But first, we'll add a helper function that handles the case of imbalanced word lists:

```{r}
# source = https://github.com/chadaeun/weat_replication/blob/master/lib/utils.py#L116C1-L133C16
balance_word_vectors <- function(A, B) {
  # Balance size of two lists of word vectors by randomly deleting some vectors in larger one.
  # If there are words that did not occur in the corpus, some words will ignored in `get_word_vectors`.
  # So result word vectors' size can be unbalanced.
  # :param A: matrix of word vectors
  # :param B: matrix of word vectors
  # :return: list of two balanced word vector matrices
  
  diff <- nrow(A) - nrow(B)

  if (diff > 0) {
    A <- A[-sample(nrow(A), diff), , drop = FALSE]
  } else if (diff < 0) {
    B <- B[-sample(nrow(B), -diff), , drop = FALSE]
  }

  return(list(A, B))
}
```

Now we are ready to compute the WEAT effect sizes for each

```{r}
res <- list()
# adapted from https://github.com/chadaeun/weat_replication/blob/master/weat_test.py#L26-82
for (data_name in names(wordlists)) {
  data_dict <- wordlists[[data_name]]
  
  if (data_dict$method == 'wefat') {
    next
  }
  
  A <- get_word_vectors(data_dict[[data_dict$A_key]])
  B <- get_word_vectors(data_dict[[data_dict$B_key]])
  
  balanced_AB <- balance_word_vectors(A, B)
  A <- balanced_AB[[1]]
  B <- balanced_AB[[2]]
  
  num_attr <- nrow(A)
  
  X <- get_word_vectors(data_dict[[data_dict$X_key]])
  Y <- get_word_vectors(data_dict[[data_dict$Y_key]])
  
  balanced_XY <- balance_word_vectors(X, Y)
  X <- balanced_XY[[1]]
  Y <- balanced_XY[[2]]
  
  num_target <- nrow(X)

  score <- weat_score(X, Y, A, B)

  res[[data_name]] <- as.data.frame(data_dict[c("targets", "attributes", "method")])
  res[[data_name]]$score <- score
  res[[data_name]]$num_target <- num_target
  res[[data_name]]$num_attr <- num_attr
}
```

Summarize the results in a Table:

```{r}
library(dplyr)
result_df <- bind_rows(res, .id = "data_name")
result_df[c("targets", "attributes", "score")]
```

Caliskan *et al.* report the following numbers in their paper:

| Targets | Attributes | WEAT | $p$-value |
|:------- |:--------- |:---- |:--------- | 
| European-American vs. African-American names	|	Pleasant vs. unpleasant	|	1.41	|	$10^{−8}$ |
| Flowers vs. insects	|	Pleasant vs. unpleasant	|	1.50	|	$10^{−7}$ |
| Male vs. female names	|	Career vs. family	|	1.81	|	$10^{−3}$ |
| Instruments vs. weapons	|	Pleasant vs. unpleasant	|	1.53	|	$10^{−7}$ |
| Science vs. arts	|	Male vs. female terms	|	1.24 | $10^{−2}$ |
<!--| Mental vs. physical disease	|	Temporary vs. permanent	|	1.38	|	$10^{−2}$ |-->
<!--| Young vs. old people's names	|	Pleasant vs. unpleasant	|	1.21	|	$10^{−2}$ |-->

So the scores we obtain from the GloVe model are mostly pretty similar.

## Computing $p$-values

What's missing so far, however, are $p$-values indicating whether our effect size point estimates are statistically significant.

So lets' get back to our example of the association of Male and Female words with Science vs. Arts words: 

```{r}
cat("A:", data_dict[[A_key]], "\n")
cat("B:", data_dict[[B_key]], "\n")
cat("X:", data_dict[[X_key]], "\n")
cat("Y:", data_dict[[Y_key]], "\n")

A <- get_word_vectors(data_dict[[A_key]])
B <- get_word_vectors(data_dict[[B_key]])
X <- get_word_vectors(data_dict[[X_key]])
Y <- get_word_vectors(data_dict[[Y_key]])
```

```{r}
# point estimate
score <- weat_differential_association(X, Y, A, B)
print(score)
```

### Implementing the permutation test

The statistical significance of the point estimate is computed with the **permutation test**.

The permutation test shuffles target words in `X` and `Y` between the two opposing categories and computes the WEAT association score $s$ for each reshuffle ("permutation").

So we'll proceed as follows:

1. we stack the matrices with word vectors for terms in *X* and *Y* on top of each other into a matrix **S**
2. we create a list of all possible permutations of the row indexes of **S**
3. we apply each list of indexes to extract rows in **S** and split it row-wise in half
4. we compute the $s$ score with the two partitions of **S** instead of **X** and **Y**
5. we compare the point estimate with the resulting WEAT association scores

```{r}
# 1. stack 
S <- rbind(X, Y)
dim(S) # <== has twice as many rows as X and Y, respectively
```

```{r}
# 2. create list of possible permutations
# note: computing all permutations is not feasible, so we sample 500 permutations
set.seed(1234)  # for reproducibility
idxs <- 1:(nrow(X) * 2)
permutations <- replicate(500, sample(idxs), simplify = FALSE)
length(permutations)
```

```{r}
# 3. iterate over permutations (reshuffled indexes), ... 
scores <- numeric(length(permutations))
for (i in seq_along(permutations)) {
  p <- permutations[[i]]
  # partition stacked matrix in "fake" X and Y matrices ...
  S_ <- S[p, , drop = FALSE]
  X_ <- S_[1:nrow(X), , drop = FALSE]
  Y_ <- S_[(nrow(X) + 1):nrow(S_), , drop = FALSE]
  # ... and compute WEAT score
  scores[i] <- weat_differential_association(X_, Y_, A, B)
}

# show distribution of scores (should average to ±0)
c(mean = mean(scores), sd = sd(scores))
```

Caliskan *et al.* note that the one-sided $p$-value is 

$$
\operatorname{Pr}_i\left[s\left(X_i, Y_i, A, B\right)>s(X, Y, A, B)\right]
$$

So we can compute this value by comparing our permutation scores with the point estimate:

```{r}
# 4. compute p-value for a given score
pvalue <- sum(scores > score) / length(scores)
# print with 10 decimal places
sprintf("%0.10f", pvalue)
```

**_Note:_** Caliskan explain that this $p$-value 'is a normalized measure of how separated the two distributions (of associations between the target and attribute) are.

Let's wrap this code in a function:

```{r}
weat_p_value <- function(X, Y, A, B, seed = 1234) {
  # Returns one-sided p-value of the permutation test for WEAT score
  # CAUTION: this function samples permutations
  # :param X: target words' vector representations
  # :param Y: target words' vector representations
  # :param A: attribute words' vector representations
  # :param B: attribute words' vector representations
  # :return: p-value (float value)
  
  score <- weat_differential_association(X, Y, A, B)
  
  # 1. stack 
  S <- rbind(X, Y)
  
  # 2. permute indices
  set.seed(seed)
  idxs <- 1:(nrow(X) * 2)
  permutations <- replicate(500, sample(idxs), simplify = FALSE)
  
  # 3. partition stacked matrix in "fake" X and Y matrices and compute WEAT scores
  scores <- sapply(permutations, function(p) {
    S_ <- S[p, , drop = FALSE]
    X_ <- S_[1:nrow(X), , drop = FALSE]
    Y_ <- S_[(nrow(X) + 1):nrow(S_), , drop = FALSE]
    weat_differential_association(X_, Y_, A, B)
  })

  # 4. compute p-value for a given score
  pvalue <- sum(scores > score) / length(scores)

  return(pvalue)
}
```

Now let's again iterate over test cases in `wordlists` but now compute the WEAT score and its associated $p$-value:

```{r}
res <- list()
for (data_name in names(wordlists)) {
  data_dict <- wordlists[[data_name]]
  
  if (data_dict$method == 'wefat') {
    next
  }
  
  A <- get_word_vectors(data_dict[[data_dict$A_key]])
  B <- get_word_vectors(data_dict[[data_dict$B_key]])
  
  balanced_AB <- balance_word_vectors(A, B)
  A <- balanced_AB[[1]]
  B <- balanced_AB[[2]]
  
  X <- get_word_vectors(data_dict[[data_dict$X_key]])
  Y <- get_word_vectors(data_dict[[data_dict$Y_key]])

  balanced_XY <- balance_word_vectors(X, Y)
  X <- balanced_XY[[1]]
  Y <- balanced_XY[[2]]
        
  res[[data_name]] <- as.data.frame(data_dict[c("targets", "attributes", "method")])
  
  res[[data_name]]$score <- weat_score(X, Y, A, B)
  res[[data_name]]$pvalue <- weat_p_value(X, Y, A, B)
  
}
```

```{r}
result_df <- bind_rows(res, .id = 'data_name')
result_df[c('targets', 'attributes', 'score', 'pvalue')]
```